---
title: "Day 3 Demo: R Notebook"
author: Benjamin Hetman
date: "2022-02-08"
output: html_notebook
---

### This is an R Markdown notebook. 

#### These notebooks are an excellent option when you're testing out new code or developing an analysis becase they make it super easy to capture your thought process and intermediate results along the way. 

The same conventions apply as *markdown documents* that you would knit to *pdf*, *word*, or *powerpoint*, however these notebooks are intended more as a *workbook format* and give you some **enhanced formatting** to allow you to follow what your code was doing later. 

Recall: refer to the [RMarkdown Cheatsheets](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) for Markdown tips and tricks! 

> Each time you save the notebook, R will render it as an ***HTML file*** that you can open and browse in your web browser. Or you can click *PREVIEW* at the top of the screen and you'll see a preview of the notebook in the RStudio Viewer pane. 

```{r setup, include=FALSE}

# Recall that because we set include=FALSE above, we can write anything we want in this box and it won't be output to the final document. 
# The code in the box will still run, however. 

library(tidyverse)
here::i_am("Day3_Demo.Rmd")
library(here)

```

## Reading Data:
As we saw in the **10 Considerations for Data Analysis**, the first step we'll often want to consider is reading our data into R, so lets do that now! 

```{r reading_data, message=FALSE}
cavity1 <- read_tsv(here("data","cavity_pt1.tsv"))
cavity2 <- read_tsv(here("data","cavity_pt2.tsv"))
desserts <- read_tsv(here("data","dessert.tsv"))

```

Let's go ahead and have a quick look at some of the data we just loaded:
```{r}
cavity1
cavity2
# print(cavity1)
# print(cavity2)
```
## Append Data
It looks like **cavity1** and **cavity2** are incomplete, we'll need to append those together to get the full set of cavity data:
```{r}
cavities <- bind_rows(cavity1, cavity2)
print(cavities)
```


## Clean Data
Hmm, *Joanne* has been duplicated in the dataset.. what are some ways that we can drop the duplicated record?

```{r}
# We could use base R indexing like we learned in Day 1, and drop the 3rd row:
# cavities <- cavities[-3,]
# print(cavities)

# Alternatively, we can use a more tidyverse approach - you'll learn to develop your own preferences over time
cavities <- cavities %>%
  distinct(.keep_all = TRUE)
print(cavities)
```
Great! That worked, now lets look at the last table we loaded named *desserts*: 
```{r}
print(desserts)
```

## Merge Data
Now we can practice our hard-earned **join** skills by applying a **full-join** to keep all records from both complete tables:
```{r}
adulting_is_hard <- 
  cavities %>% 
  full_join(desserts)
  
print(adulting_is_hard)
```

## Data Visualization
Finally, this wouldn't be a proper R demonstration without some kind of ***fun visualization***: 
```{r Visualization, echo=FALSE, fig.height=5, fig.width=6}
library(ggimage)

images <- c("images/icedcoffee.png",
            "images/croissant.png",
            "images/tiramisu.png",
            "images/chocolate.png",
            "images/cappacino.png",
            "images/peanutbutter.png",
            "images/gummybear.png",
            "images/orange_cake.png"
            )

adulting_is_hard$image <- images

ggplot(data = adulting_is_hard, aes(x=Desserts.wk, y=Cavities.10yr)) +
  geom_image(aes(image=image), size = 0.1) + 
  geom_label(aes(label = Name), nudge_x = 0.5, nudge_y = 0.25) + 
  theme_minimal() +
  labs(title="Let them eat cake!",
       subtitle="(But also let's promote good dental hygiene)",
       x = "Avg Desserts per Week",
       y = "Total # Cavities in 10 year period: 2010-2019")

```

And if you want to start getting fancy, check out the [plotly](https://plotly.com/r/animations/#frames) package in R!! 
```{r plotly, message=FALSE}

library(plotly)

cavity_year <- read_tsv(here("data", "cavities_yr.tsv")) %>% 
  left_join(adulting_is_hard) %>% 
  pivot_longer(2:11, names_to="Year")

cavity_fig <- cavity_year %>% 
  plot_ly(
    x =~Name, 
    y= ~value,
    frame= ~Year,
    type="scatter", 
    mode="markers",
    showlegend=F,
    color = ~Name
    )

cavity_fig
```


