---
title: "Day3_Exercise_Script"
output: html_document
---

##### Script Information: 
*Purpose:* 
Exercise in advanced data management training for epidemiologists in R by using mock admin datasets to identify cases of Asthma
         
*Author:*
J. Stares
*Date:*
26-Oct-2020

*Last modified by:*
M.A. Belair
B. Hetman

*Date last modified: *
2021-11-17

##### Notes: 
* Mock admin SAS datasets processed in R and Stata in 2017
* Training module location: Dropbox
* Case definition - Asthma: 1 year and older - 1H Ever or 2P in 2Y https://www.canada.ca/en/public-health/services/publications/canadian-chronic-disease-surveillance-system-factsheet.html


### Setup
```{r setup, include=FALSE}
require("knitr")
library(tidyverse)
library(here)

knitr::opts_chunk$set(echo = TRUE)
here::i_am("Exercise_Day3/scripts/Exercise_Day3.Rmd")
output_folder <- here::here("Exercise_Day3", "output")

```

## R Markdown

##### Preparing datasets
##### DAD

```{r}
dad1213 <- read_csv(here("Exercise_Day3", "data", "dad1213.csv"))
dad1314 <- read_csv(here("Exercise_Day3", "data", "dad1314.csv"))
```

##### Append DAD datasets:

```{r}
dad_appnd <- bind_rows(dad1213, dad1314) %>%
  mutate(sepdat = as.Date(sepdat, "%d-%b-%Y")) %>%
  filter(sepdat > "2012-04-01" & sepdat < "2014-03-31")

```

##### Process DAD Datasets:
```{r}
#create flag variable.  
####Step 1: Extract first three digits of diag code.
dad_appnd <- dad_appnd %>% mutate(diag1_sbstr = str_sub(diag1, end=3)) %>%
  mutate(diag2_sbstr = str_sub(diag2, end=3)) %>%
  mutate(diag3_sbstr = str_sub(diag3, end=3))

####Step 2: create flag where 0 = no asthma, 1= asthma if any diag variables equal to 493.
dad_appnd <- dad_appnd %>% mutate(flag = case_when(diag1_sbstr == "493" | diag2_sbstr =="493" | diag3_sbstr =="493" ~ 1, TRUE ~ 0))



#keep unique asthma-related hosps
#####Step 1: Keep only asthma related events.
#####Step 2: Group by unique ID (a) and count how many uniqueIDs appear in each group (b)
#####Step 3: Keep only earliest sepdat where  unique persons are duplicated.
#####Step 4: Keep (a) and rename (b) only the variables you need for your analysis.
dad_appnd_asthma <- dad_appnd %>% 
  filter(dad_appnd$flag == 1) %>% #Step 1
  group_by(uniqueid) %>% #Step 2 (a)
  add_tally(name = "count_duplicates") %>% #Step 2 (b)
  arrange(uniqueid, sepdat) %>% #Step 3 (a)
  distinct(uniqueid, .keep_all = TRUE) %>% #Step 3 (b)
  select(uniqueid, sepdat, count_duplicates) %>% #Step 4 (a)
  rename(dad_sepdat_asthma = sepdat, dad_count_asthma=count_duplicates) #Step 4 (b)

#tidy workspace
rm(dad_appnd, dad1213,dad1314)

```

#### Prepare MSP datasets
```{r}
##load files
msp1213 <- read_csv(here("Exercise_Day3", "data", "msp1213.csv"))
msp1314 <- read_csv(here("Exercise_Day3", "data","msp1314.csv"))

##append datasets
msp_appnd <- bind_rows(msp1213, msp1314)

### process MSP.
########Step 1 - Reformat servdat to date-class variable.
########Step 2 - Filter by servdat: keep if servdat between Apr 1 2012 & Mar 13 2014
########Step 3 - Extract first three digits from diag1.
########Step 4 - Create new flag variable, 1=asthma, 0=no asthma; code block 493 (incl 4390...4939)

msp_appnd <- msp_appnd %>%
  mutate(servdat = as.Date(servdat, "%d%b%Y")) %>% #Step 1
  filter(servdat >= "2012-04-01" & servdat <= "2014-03-31") %>% #Step 2
  mutate(diag1_sbstr = as.numeric(str_sub(diag1, end=3))) %>% #Step 3
  mutate(flag = case_when(diag1_sbstr == "493" ~ 1, TRUE ~ 0)) #Step 4


#Step 1 - Keep & count phys visits for asthma
#Step 2 - Identify and count how many observations per uniqueid
#Step 3 - keep only earliest servdat where unique persons are duplicated
msp_appnd_asthma <- msp_appnd %>% filter(msp_appnd$flag == 1) %>% #Step 1
  group_by(uniqueid) %>% add_tally(name="count_duplicates") %>% #Step 2
  arrange(uniqueid, servdat) %>% distinct(uniqueid, .keep_all = TRUE) #Step 3

#Tidy the data.
msp_appnd_asthma <- msp_appnd_asthma %>%
  select(uniqueid, servdat, count_duplicates) %>%
  rename(msp_servdat_asthma = servdat, msp_count_asthma = count_duplicates)

#clear work environment.
rm(msp_appnd, msp1213, msp1314)
```


##### Assemble datasets with registry file

###### Registry Dataset
```{r}
#####Load registry file
reg1314 <- read_csv(here("Exercise_Day3", "data", "reg1314.csv"))


#####Merge dad to registry
asthma_dad_reg1314 <- full_join(reg1314, dad_appnd_asthma, by=c("uniqueid"="uniqueid"))

##freq of each result in dad_count_asthma
asthma_dad_reg1314 %>% count(dad_count_asthma)

#####Merge MSP to registry
asthma_msp_reg1314 <- full_join(reg1314, msp_appnd_asthma, by=c("uniqueid"="uniqueid"))

asthma_msp_reg1314 %>% count(msp_count_asthma)

```

###### Prepare Asthma-Registry datasets
```{r}
#create combined asthma registry datasets.
asthma_reg1314 <- full_join(asthma_dad_reg1314, asthma_msp_reg1314, by=c("uniqueid"="uniqueid", "sex"="sex", "dob"="dob"))

table(asthma_reg1314$msp_count_asthma, asthma_dad_reg1314$dad_count_asthma)
count(asthma_reg1314, msp_count_asthma)
count(asthma_dad_reg1314, dad_count_asthma)
```

###### Process Asthma-Registry Dataset
```{r Flag individuals who meet hospitalisation and physician visit portion of the case definition}

# 1H Ever or 2P in 2Y
asthma_reg1314 <- asthma_reg1314 %>%
  mutate(PH_flag = case_when(dad_count_asthma >= 1 ~ 1,
                             msp_count_asthma >= 2 ~ 1,
                             TRUE ~ 0))


asthma_reg1314 %>% count(PH_flag)
```

```{r Calculate Age}
asthma_reg1314 <- asthma_reg1314 %>% mutate(dob=as.Date(dob, "%d%b%Y")) %>%
  mutate(age = as.numeric(difftime(time1 = as.Date("01jul2013", "%d%b%Y"), time2 = dob, units="days"))/365.25)

```


```{r Identify asthma cases based on age and case definition}
#143 cases
asthma_reg1314 <- asthma_reg1314 %>% 
  mutate(popn_1plus = case_when(age >= 1 ~ 1, TRUE ~ 0)) %>%
  mutate(asthma_case = case_when(popn_1plus == 1 & PH_flag == 1 ~ 1, TRUE ~ 0))

asthma_reg1314 %>% count(asthma_case)
```

```{r Process when case met definition}
asthma_reg1314 <- asthma_reg1314 %>%
  mutate(date_flag = case_when(
    asthma_case == 1 & dad_sepdat_asthma < msp_servdat_asthma ~ 1,
    asthma_case == 1 & dad_sepdat_asthma > msp_servdat_asthma ~ 2,
    asthma_case == 1 & dad_sepdat_asthma == msp_servdat_asthma ~ 3,
    asthma_case == 1 & is.na(dad_sepdat_asthma) == TRUE ~ 4,
    asthma_case == 1 & is.na(msp_servdat_asthma) == TRUE ~ 5,
    TRUE ~ 0))

asthma_reg1314 <- asthma_reg1314 %>%
  mutate(asthma_casedate = case_when(
    date_flag == 1 ~ dad_sepdat_asthma,
    date_flag == 2 ~ msp_servdat_asthma,
    date_flag == 3 ~ msp_servdat_asthma,
    date_flag == 4 ~ msp_servdat_asthma,
    date_flag == 5 ~ dad_sepdat_asthma))


#count asthma case flags
asthma_reg1314 %>% count(asthma_case)

#count date flags
asthma_reg1314 %>% count(date_flag)

#compare case and date flags
asthma_reg1314 %>% count(asthma_case, date_flag)

#compare case date and case flags
asthma_reg1314 %>% count(asthma_casedate,date_flag)

#Tidy Data
str(asthma_reg1314)

#select variables to keep
asthma_reg1314 <- asthma_reg1314 %>%
  select(uniqueid, dob, age, sex, asthma_case, asthma_casedate, popn_1plus)

#clean up work environment
rm(asthma_dad_reg1314, asthma_msp_reg1314, dad_appnd_asthma, msp_appnd_asthma, reg1314)
```





###### Data Analysis

```{r Calculate crude prevalence}
#crude prevalence:
#add up n cases and n population


asthma_overall <- asthma_reg1314 %>%
  summarize(cases = sum(asthma_case), popn = sum(popn_1plus)) %>%
  mutate(crude_rate = (cases/popn)*100000)

print(asthma_overall)
```

```{r Calculate age-grouped crude prevalence}
asthma_reg1314 <- asthma_reg1314 %>%
  mutate(agegrp = case_when(
    age >= 1 & age <= 9 ~ "<10",
    age >= 10 & age <= 19 ~ "10-19",
    age >= 20 & age <= 29 ~ "20-29",
    age >= 30 & age <= 39 ~ "30-39",
    age >= 40 & age <= 49 ~ "40-49",
    age >= 50 & age <= 59 ~ "50-59",
    age >= 60 & age <= 69 ~ "60-69",
    age >= 70 & age <= 79 ~ "70-79",
    age >= 80 & age <= 89 ~ "80-89",
    age >= 90 ~ "90+" ))

#tabulate age groupings
asthma_reg1314 %>% count(agegrp)

#tabulate cases per age grouping
asthma_reg1314 %>%
  count(agegrp, asthma_case) %>%
  pivot_wider(names_from = asthma_case, values_from = n)

#Summarize cases and population per age range

asthma_agespecific <- asthma_reg1314 %>%
    group_by(agegrp) %>%
  summarize(cases = sum(asthma_case), popn = sum(popn_1plus), .groups = "drop") %>%
  filter(!is.na(agegrp))


#calculate rates
asthma_agespecific <- asthma_agespecific %>%
  mutate(agespecific_rate = (cases/popn)*1000000)

```


#Age Specific Plot
```{r Basic Age Specific Plot}
#basic plot:

agespec_barchart <- ggplot(data=asthma_agespecific, aes(x=agegrp, y=agespecific_rate)) +
  geom_bar(stat="identity")

#add simple options to the plot

agespec_barchart <- agespec_barchart +
  labs(title="Age-Specific Asthma Rate", x="Age Group (10 Year)", y="Rate (per 100,000)")+
  theme_classic()

#call plot
agespec_barchart
```

```{r Age and sex-specific asthma rates}

#add sex variable to the dataset

asthma_reg1314 <- asthma_reg1314 %>%
  mutate(sexchar = case_when(asthma_reg1314$sex == 1 ~ "Female",
                             asthma_reg1314$sex == 2 ~ "Male",
                             TRUE~"Unknown"))

#summarized based on age and sex

asthma_as_specific <- asthma_reg1314 %>%
  group_by(agegrp, sexchar) %>%
  summarize(cases = sum(asthma_case), popn = sum(popn_1plus), .groups = "drop") %>%
  filter(!is.na(agegrp))

#calculate rates
asthma_as_specific <- asthma_as_specific %>%
  mutate(as_specific_rate = (cases/popn)*100000)


#plot age and sex-specific results

agesexspec_barchart <- ggplot(data=asthma_as_specific, aes(fill=sexchar, x=agegrp, y=as_specific_rate))+
  geom_bar(position="dodge", stat="identity")+
  labs(title="Age- and Sex-Specific Asthma Rate",
      x="Age Group (10 year)",
      y="Rate (per 100,000)",
      fill="Sex") +
  theme_classic()

#call plot
agesexspec_barchart

```

## BONUS MATERIAL
## Note: the following section is completely optional, and included to demonstrate some more advanced options available for your data-processing and visualization in R. 


##### So many plotting options!  
```{r Demonstrating a custom function}
# Create your own function to format plot styles! 

theme_awesome <- function(){
  theme_classic() +
    theme(
      line = element_line(color="black"),
      panel.background = element_rect(fill = "white", colour = "darkgrey"),
      panel.grid.major.y =  element_line(colour = "grey", linetype = 3, size = 0.5),
      plot.background = element_rect(fill="white"),
      strip.background = element_rect(fill = "white"), 
      legend.background = element_rect(fill = "white"),
      legend.key = element_rect(fill = "white"), 
      legend.position = "bottom",
      plot.title = element_text(size = 14, family = "sans", face = "bold"),
      plot.subtitle = element_text(size = 9, family = "sans", face = "italic"),
      axis.title.x = element_text(),
      text = element_text(size = 11), 
      
  )
}

agesexspec_barchart + 
  theme_awesome() + 
  scale_fill_manual(values = c("#4D648D", "#D0E1F9")) + 
  scale_color_manual(values = "darkgrey") +
  labs(subtitle = "Great job for getting this far. Enjoy some cake!")

```

##### Demonstrating facets and a simple loop:
```{r Facetting on sex}
# What if we wanted to compare sexes by themselves?
agesexspec_barchart + 
  theme_awesome() + 
  scale_fill_manual(values = c("#4D648D", "#D0E1F9")) + 
  scale_color_manual(values = "darkgrey") +
  labs(subtitle = "Great job for getting this far. Enjoy some cake!")+
  facet_wrap(~sexchar)

```
```{r A simple loop}
# What if we wanted a separate chart for each age group?
# The following code will iterate through each age group created earlier, 
# and produce a unique chart for that age category only; then save that chart 
# in your output directory.


for (i in unique(asthma_as_specific$agegrp)){
# Create titles for the individual plots based on the age group
title <- paste0("Asthma rate among age group ", i)
# Special note: R will fail if the file name includes non-alpha-numeric characters
# We use str_replace_all to remove these
title <- str_replace_all(title, "[^[:alnum:]]", " ")

#subset the data we need to plot
loop_plot <- asthma_as_specific %>% filter(agegrp==i)

# Now we plot and save the individual files to our working directory
ggplot(data=loop_plot, aes(fill=sexchar, x=agegrp, y=as_specific_rate))+
  geom_bar(position="dodge", stat="identity")+
  labs(title="Age- and Sex-Specific Asthma Rate",
      x=paste0("Age (years)"),
      y="Rate (per 100,000)",
      fill="Sex") +
  theme_awesome() + 
  scale_fill_manual(values = c("#4D648D", "#D0E1F9")) + 
  scale_color_manual(values = "darkgrey") +
  labs(title = title, 
       subtitle = "Great job for getting this far. Enjoy some cake!")

ggsave(path = output_folder, filename = paste0(title, ".png"), device = "png")
}
#Check your output directory for the exported plots! 

```

## Bonus Challenge! 
Now that you have your plots saved as outputs, revisit the Day2_final.Rmd script 
and see if you can use R-Markdown to include these as embedded images. 

Hint: try using the knitr::include_graphics() function! 


